---
title: "Fibonacci Spirals"
output:
  html_document:
    code_folding: hide
    
---
```{r setup, echo = FALSE, warning=FALSE}
library(ggplot2)
library(gganimate)
```
I have always found a sense of peace through painting and drawing geometric patterns. After all, geometry is the language of nature. I draw my inspiration from the ancient and medieval geometric art to recreate the patterns using a ruler and compass on a tracing paper, then transfer to the painting paper(watercolor). This time though, I wanted to do this using a computer program. This project is one example pattern created using the Fibonacci spirals. 

Before writing this post, I wasn't familiar with the term Fibonacci spirals. I was only looking to find a way to draw some of my geometric patterns in R. I stumbled upon a picture of the center of a sunflower, showing the beautiful arrangement of the seeds in the middle.And I wondered, if I could create that pattern...

[This](https://blog.wolfram.com/2011/07/28/how-i-made-wine-glasses-from-sunflowers/) post helped towards understanding the mathematics behind the arrangement. Three things noted in the arrangement are


* The seeds spiral outward, with each seed positioned at a fixed angle relative to its predecessor.


* The fixed angle is given by $\Gamma = 2\pi(1-\phi)$, where $\phi = 1.61803...$ is the golden ratio.


* The ith seed is placed at a radial distance of $\sqrt(i)$


Hence, in the polar coordinates, we can specify $[r,\theta] = [\sqrt(i), \Gamma*i]$. Since I wanted to plot the points on cartersian coordiantes, I used the polar to Cartesian conversion factors: 


1. $x = rCos(\theta)$ 


2. $y = rSin(\theta)$
and plugged in the values of r and the fixed angle $\Gamma$


```{r}
x<- c()   
y <- c()  

goldenratio <- 1.618
gamma <- 2*pi*(1-goldenratio)  

for (i in 1:1000){
  x<- c(x,sqrt(i)*cos(i*gamma))
  y<- c(y,sqrt(i)*sin(i*gamma))
 
}

df <- data.frame(x = x, y = y)

p <- ggplot(data = df, aes(x=x,y=y))+
  geom_point()

p
```

I generated the animation [here](index.html) using gganimate() package. I experimented with values of $\phi$ other than golden ratio in the graphic. 

