---
title: "Projects"
output: html_document
---

This is a dummy project. I picked a data set from here: 

The Growth Lab at Harvard University, 2019, "Growth Projections and Complexity Rankings", https://doi.org/10.7910/DVN/XTAQMC, Harvard Dataverse, V2, UNF:6:+dXp8TMQz26OFv7ZOfIxSg== [fileUNF]

And country to continent data from here:
https://www.kaggle.com/statchaitya/country-to-continent?select=countryContinent.csv

I do not understand the science behind this data set.For now -  The goal is to be able to produce and publish some graphics, using R markdown


```{r}
rankingsData <- read.delim("Data/rankings.tab")
locationData <- read.delim("Data/location.tab")

df <- merge(x = rankingsData, y = locationData, by = "location_id", all = TRUE)


continentdt <- read.csv("Data/continent.csv")
colnames(continentdt)[3] <- "code"
continentclean <- subset(continentdt, select= c(country,code, continent))

df2 <- merge(x = df, y = continentclean, by = "code", all = TRUE)


#plot 1
library(ggplot2)
qplot(x=hs_eci, y=sitc_eci, 
      data=df, 
      colour=year, 
      main="HS ECI vs SITC ECI")

#plot 2
ggplot(data = df2) +
  geom_point(mapping= aes(hs_eci, growth_proj, 
                          colour = continent))+
  ggtitle("Country data grouped by continent")


```

The first R code chunk ends here. Next chunk:
```{r}
#plot3
ggplot(data = df2) +
  geom_point(mapping= aes(hs_eci, growth_proj, 
                          colour = continent))+
  facet_wrap(~ continent, nrow = 3)+
  ggtitle("Facet Wrap by continent")
  
#plot 4
  ggplot(data = df2, mapping= aes(x= hs_eci,y = growth_proj)) +
geom_point(mapping= aes(color = continent))+
geom_smooth()+
ggtitle("Smooth line with points")
```


